A comparison of R tools for nonlinear least squares modeling

### 2023-05-11 RefactoringNLS/ComparingNLStools.Rmd
JN final edited and reduced

`onls` (@R-onls-manual) -- The usual optimization in estimating nonlinear models
is the vertical difference between the dependent variable and the functional
model. `onls` minimizes the sum of **orthogonal** residuals. The objective is
therefore different and involves nontrivial extra calculation. A vignette with
the package and the blog article 
https://www.r-bloggers.com/2015/01/introducing-orthogonal-nonlinear-least-squares-regression-in-r/ 
give some description with illustrative graphs. `onls` appears to be limited to 
problems with one independent and one dependent variable. The Wikipedia article
https://en.wikipedia.org/wiki/Total_least_squares presents an overview of some
ideas, with references to the literature. Our view is that the approach needs wider 
discussion and tutorial examples to allow its merits to be judged.

`crsnls` (@R-crsnls-manual) -- This package allows nonlinear estimation by 
controlled random search via two methods. There is unfortunately no vignette.
A modest trial we carried out showed `nlsr::nlxb()` gave the same results in 
a small fraction of the time required by either of the methods in `crsnls`. The
method discussed in @tvrdik2007 claims better reliability in finding solutions
than a Levenberg-Marquardt code (actually from Matlab), but the tests were conducted
on the extreme NIST examples (@NISTnls-manual).

@@

`nlsic` (@nlsic-manual) -- This solves nonlinear least squares problems with optional
equality and/or inequality constraints. It is clearly **not** about modeling, and
the input and output are quite different from class `nls` methods.

`nlsMicrobio` (@nlsMicrobio-manual) -- Data sets and nonlinear regression models 
dedicated  to predictive microbiology, including a vignette, by authors of the
`nlstools` package who have a strong record in the subject.

@@

`nls2` (@nls2-22) -- Nonlinear least squares by brute force has similar
motivations, but uses `nls()` within multiple trials.

@@

`nlraa`  (@MiguezNLRAA2021) -- a set of nonlinear *selfStart* models. Most include
analytic Jacobian code.

@@

The functions above are equivalent, but the first is generally more awkward
to solve numerically due to its poor scaling. The parameters of the three forms
are related as follows:

@@

Let us consider there are $n$ parameters and $m$ residuals. 
We note several useful quantities. The loss function is 
$$ S(p) = r' r = \sum_{i=1}^m { r_i^2 }$$
The gradient of $S(p)$ is
$$ 2 * J' r$$
where the Jacobian $J$ is given by elements
$$ J_{i,j} = \partial r_i / \partial p_j $$
and the Hessian is defined by elements
$$ H_{i,j} = \partial ^2 S(p) / {\partial p_i \partial p_j} $$
If we expand the Hessian for nonlinear least squares problems, we find
$$  H_{i,j} = \sum_{k = 1}^m{ J_{k,i} J_{k,j}}  + \sum_{k = 1}^m {r_k * \partial{r_k}/{\partial p_i \partial p_j}}$$
Let us use $D_{i,j}$ for the elements of the second term of this expression.
What is generally called **Newton's method** for function minimization tries
to set the gradient to zero (to find a stationary point of the function $S(p)$).
This leads to **Newton's equation**
$$ H \delta = -g $$
Given a set of parameters $p$, we solve this equation for $\delta$, adjust $p$
to $p + \delta$ and iterate, hopefully to converge on a solution. 
Applying this to a sum of squares problem gives
$$ H \delta = (J'J + D) \delta = - J' r$$
In this expression, only the elements of $D$ have second partial derivatives. 
Gauss, attempting to model planetary orbits, had small residuals, and noted
that these multiplied the second partial derivatives of $r$, so he approximated
$$ H \approx J' J$$
by assuming $D \approx 0$. This results in the Gauss-Newton method where we
solve 
$$ J' J \delta = - J' r$$
though we can avoid some loss of accuracy by NOT forming the inner product matrix
$J' J$ and solving the linear least squares matrix problem 
$$ J \delta \approx -r $$
by one of several matrix decomposition methods.

@@

The usual suggestion is that $D$ be replaced by a multiple of the unit matrix or else
a multiple of the diagonal part of $J' J$. In low precision, some elements of 
$J' J$ could underflow to zero (@jn77ima), and one of us suggested a linear combination of both
choices as a compromise. Various choices for
$D$, as well as a possible line search along the
direction $\delta$ rather than a unit step (@Hartley1961), 
give rise to several variant
algorithms. "Marquardt's method" is a family of methods. Fortunately, most choices
work well.

@@

Let us specify in R the three model formulas and set some starting
values for parameters. These starting points are NOT equivalent and
are deliberately crude choices. Workers performing many calculations 
of a similar nature should try to provide good starting points to reduce
computation time and the chance a false solution is found.

@@

Using a formula specification was a principal advantage made
with `nls()` when it became available in S sometime in the 1980s. It uses a
Gauss-Newton (i.e., unstabilized) iteration with a step reduction line
search. This works very efficiently as long as $J$ is not ill-conditioned.
Below we see `nls()` does poorly on the example problem. To save page space,
we use 1-line result display functions from package `nlsr`, nameley `pnlm0()` 
and `pshort()`.

@@

```{r ex02minpack}
unlm1<-try(nlsLM(formula=frmu, start=stu1, data=weeddf))
pnlm0(unlm1)  # Short form of output
snlm1<-try(nlsLM(formula=frms, start=sts1, data=weeddf))
pnlm0(snlm1)
tnlm1<-try(nlsLM(formula=frmt, start=stt1, data=weeddf))
pnlm0(tnlm1) # short form to give sum of squares, else use summary(tnlm1)
```

@@

```{r ex02gslnls}
library(gslnls)
ugslnls1<-try(gsl_nls(fn = frmu, data = weeddf,  start = stu1))
pnlm0(ugslnls1) # to get sum of squares
sgslnls1<-try(gsl_nls(fn = frms, data = weeddf,  start = sts1))
pnlm0(sgslnls1) # Use summary() to get display
tgslnls1<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1))
pnlm0(tgslnls1) 
```

### Comparison notes for formula setup solutions

@@

We illustrate how to solve nonlinear least squares problems using
a function to define the residual. Note that `gsl_nls()` requires
a vector `y` that is the expected value of what we have called the
residual, but in this case is actually the model. `gsl_nls` uses 
a numerical approximation for the Jacobian if the argument `jac` 
is missing.

```{r exhobbsfn, eval=TRUE}
hobnlfb<-nlfb(start=stu1, resfn=hobbs.res, jacfn=hobbs.jac)
pshort(hobnlfb) # use print(hobnlfb) for more detail
hobnlm<-nls.lm(par=stu1, fn=hobbs.res, jac=hobbs.jac)
pnlm1(hobnlm)  
hobgsln<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12))
pnlm0(hobgsln)
hobgsl<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12), jac=hobbs.jac)
pnlm0(hobgsl) # using analytic Jacobian
```

@@

The output of `nls()` is an object of class "nls" which has quite rich structure
described in the manual file or revealed by applying the `str()` function to the
result of `nls()`.

The complexity of this object is a challenge to users. Let us use 
`result` as the returned object from `nls()`, `minpack.lm::nlsLM()` or
`gslnls::gsl_nls()`. For example, the `data` return element is an R symbol. To
actually access the data from this element, we need to use the syntax:
```
eval(parse(text=result$data))
```
However, if the call is made with `model=TRUE`, then there is a returned element
`model` which contains the data, and we can list its contents using:
```
ls(result$model)
```
and if there is an element called `xdata` that can be accessed as
`result$model$xdata`.

@@

Using the "gradient" attribute of the output of the Jacobian function to hold
the Jacobian matrix lets us embed this in the **residual** function as well,
so that the call to `nlfb()` can be made with the same name used for 
both residual and Jacobian function arguments. This programming trick saves a 
lot of trouble for the package developer, but it can be a nuisance for users 
trying to understand the code.

@@

The `nls()` function, after
computing the "right hand side" or `rhs` of the residual, checks to see if the
"gradient" attribute is defined, and, otherwise using `numericDeriv()` to compute a 
Jacobian into that attribute. This code is within the `nlsModel()` or
`nlsModel.plinear()` functions. The use of analytic Jacobians
almost certainly contributes to the good performance of `nls()` on `selfStart` 
models.

@@

In considering `selfStart` models, we noted that the base-R function `SSlogis`
is intended to solve problem **Logistic3T** above.  When this function is used
via `getInitial()` to  find
starting values, it actually calls `nls()` with the 'plinear' algorithm 
and finds a solution. It then passes the solution coefficients to the default 
algorithm unnecessarily. Moreover, the double call is, in our view, prone to creating 
errors in code maintenance. 
To provide simpler starting parameters, the function `SSlogisJN`
is now part of package `nlsr`, but is most useful for `nls()`.

Users may also want to profit from the Jacobian code of `selfStart` models 
but supply explicit starting values other than those suggested by `getInitial()`. 
This does not appear to be possible with `nls()`. `nlsr::nlxb()` always requires 
starting parameters, and uses `getInitial()' to find them from the selfStart
model. The same model is used in the formula for the `nlxb()` call.

@@

We are also surprised that the 
analytic expressions for the Jacobian ("gradient") in the `SSLogis` R function 
and others save quantities in "hidden"
variables, i.e., with names preceded by ".". 
These are then not displayed by the `ls()`
command, making them difficult to access by users who may wish to create
their own selfStart model by a copy and edit.
Interactive tools, such as "visual fitting" (@nash1996nonlinear) might be
worth considering as another way to find starting parameters, but we know of no
R capability of this type.

@@

In some cases, we know that parameters cannot be bigger or smaller than some
externally known limits. Such limits should be built into models, but there 
are some important details for using the tools in R.

@@

```{r ex10, echo=TRUE} 
# Start MUST be feasible i.e. on or within bounds
anlshob1b <- nls(frms, start=sts1, data=weeddf, lower=c(0,0,0),
             upper=c(2,6,3), algorithm='port')
pnls0(anlshob1b) #  check the answer (short form)
# nlsLM seems NOT to work with bounds in this example
anlsLM1b <- nlsLM(frms, start=sts1, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
pnls0(anlsLM1b)
# also no warning if starting out of bounds, but gets good answer!!
st4<-c(c1=4, c2=4, c3=4)
anlsLMob <- nlsLM(frms, start=st4, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
pnls0(anlsLMob)
# Try nlsr::nlxb()
anlx1b <- nlxb(frms, start=sts1, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
pshort(anlx1b)
```

@@

```{r ex10m, echo=TRUE}
# Hobbsmaskx.R -- masks with formula specification of problem
require(nlsr); require(minpack.lm); traceval<-FALSE
stu <- c(b1=200, b2=50, b3=0.3) # a default starting vector (named!)
sts <- c(c1=2, c2=5, c3=3) # a default scaled starting vector (named!)
# fix first parameter
anxbmsk1 <- try(nlxb(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
			upper=c(200, 60, 3), trace=traceval))
print(anxbmsk1)
anlM1 <- try(nlsLM(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
			upper=c(200, 60, 3), trace=traceval))
pnlm0(anlM1)
anlsmsk1 <- try(nls(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
		upper=c(200, 60, 3),  algorithm="port", trace=traceval))
pnls0(anlsmsk1)
# Hobbs scaled problem with bounds, formula specification
anlxmsks1 <- nlxb(frms, start=sts, data=weeddf, lower=c(2,0,0),
                  upper=c(2,6,30))
print(anlxmsks1)
anlshmsk1 <- nls(frms, start=sts, trace=traceval, data=weeddf, lower=c(2,0,0),
             upper=c(2,6,30), algorithm='port')
pnls0(anlshmsk1)
anlsLMmsks1 <- nlsLM(frms, start=sts, data=weeddf, lower=c(2,0,0),
                 upper=c(2,6,30))
pnlm0(anlsLMmsks1)

@@

`nls()` can be called without specifying the `data` argument. In this case, it will
search in the available environments (i.e., workspaces) for suitable data objects. 
We do NOT like this approach, but it is "the R way". R allows users to leave many 
objects in the default (`.GlobalEnv`) workspace. Moreover, users have to actively 
suppress saving this workspace (`.RData`) on exit, otherwise any such file in 
the path will be loaded on startup. R users in our
acquaintance avoid saving the workspace because lurking data and
functions which may cause unwanted results.
Program alternatives to `nls()` still need to ensure they work with data in 
the available environments.

@@

`na.action` is an argument to the `nls()` function, but it does not appear 
obviously in the source code, often being handled behind the scenes after 
referencing the option `na.action`, yet this feature also changes the data 
supplied to our nonlinear least squares solver. 
A useful, but possibly dated, description is given in:
*https://stats.idre.ucla.edu/r/faq/how-does-r-handle-missing-values/*.
The typical default action, which can be seen by using the command
`getOption("na.action")`
is `na.omit`. This option removes from computations any observations
containing missing values (i.e. any row of a data frame containing an NA). 
`na.exclude` does much of the same for solver computations, but keeps the rows with 
NA elements so that predictions are in the correct row position. We recommend that 
workers actually test output to verify the behaviour is as wanted.
See *https://stats.stackexchange.com/questions/492955/should-i-use-na-omit-or-na-exclude-in-a-linear-model-in-r*.
As with `subset`, our concern with `na.action` is that users may be unaware of the
effects of an option they may not even be aware has been set. Should `na.fail` be the default?

@@

While fixed weights may be useful, there are many problems for which
we want weights that are determined at least partially from the model 
parameters, for
example, a measure of the standard deviation of observations.
Such dynamic weighting situations are discussed in the vignette 
"Introduction to nlsr" of package `nlsr` in section 
*Weights that are functions of the model parameters* and 
`minpack.lm` offers a function `wfct()` to 
facilitate such weighting. Care is advised
in applying such ideas.

@@

The variable projection method (@Golub1973, @OlearyRust13) is 
generally much more effective than general approaches in finding good solutions 
to nonlinear least squares 
problems when some of the parameters appear linearly. In our logistic examples, 
the asymptote parameters are an illustration. However, identifying which parameters
are linear and communicating this information to estimating functions 
is not a trivial task.

`nls()` has an option `algorithm="plinear"` that allows some partially linear
models to be solved. The other tools, as far as we are aware, do not offer any
such capability. The `nlstac` package uses a different algorithm for similar goals.

@@

Maintainers of packages need suitable tests and use-case examples in order
- to ensure packages work properly, in particular, giving
results comparable to or better than the functions they are to replace. 
- to test individual solver functions to ensure they work across the range of 
  calling mechanisms, that is, different ways of supplying inputs to the solver(s);
- to pose "silly" inputs to see if these bad inputs are caught by the programs.

>>>>

`onls` (@R-onls-manual) -- The usual optimization in estimating nonlinear models
is the vertical difference between the dependent variable and the functional
model. `onls` minimizes the sum of **orthogonal** residuals. The objective is
therefore different and involves nontrivial extra calculation. A vignette with
the package and the blog article 
https://www.r-bloggers.com/2015/01/introducing-orthogonal-nonlinear-least-squares-regression-in-r/ 
give some description with illustrative graphs. `onls` appears to be limited to 
problems with one independent and one dependent variable. The Wikipedia article
https://en.wikipedia.org/wiki/Total_least_squares presents an overview of some
ideas, with references to the literature. The approach needs a wider 
discussion and tutorial examples to allow its merits to be judged than can be
included here.

`crsnls` (@R-crsnls-manual) -- This package allows nonlinear estimation by 
controlled random search via two methods. There is unfortunately no vignette.
A modest trial we carried out showed `nlsr::nlxb()` gave the same results in 
a small fraction of the time required by either of the methods in `crsnls`. The
method discussed in @tvrdik2007 claims better reliability in finding solutions
than a Levenberg-Marquardt code (actually from Matlab), but the tests were conducted
on the extreme NIST examples mentioned next.

@@

`nlsic` (@nlsic-manual) -- This solves nonlinear least squares problems with optional
equality and/or inequality constraints. It is clearly **not** about modeling, and
the input and output are quite different from class `nls` methods. However, there do
not appear to be other R packages with these capabilities.

`nlsMicrobio` (@nlsMicrobio-manual) -- Data sets and nonlinear regression models 
dedicated  to predictive microbiology, including a vignette, by authors of the
`nlstools` package.

@@

`nls2` (@nls2-22) -- Nonlinear least squares by brute force has similar
motivations to `nls.multstart`, but uses `nls()` within multiple trials. The author
has extensive expertise in R.

@@

`nlraa`  (@MiguezNLRAA2021) -- a set of nonlinear *selfStart* models, primarily from
agriculture. Most include analytic Jacobian code.

@@

where we will use `weed` for $y$ and `tt` for $t$. 
The functions above are equivalent, but the first is generally more awkward
to solve numerically due to its poor scaling. The parameters of the three forms
are related as follows:

@@

Let us consider there are $n$ parameters and $m$ residuals. 
The loss function is 
$$ S(p) = r' r = \sum_{i=1}^m { r_i^2 }$$
The gradient of $S(p)$ is
$$ 2 * J' r$$
where the Jacobian $J$ is given by elements
$$ J_{i,j} = \partial r_i / \partial p_j $$
and the Hessian is defined by elements
$$ H_{i,j} = \partial ^2 S(p) / {\partial p_i \partial p_j} $$
If we expand the Hessian for nonlinear least squares problems, we find
$$  H_{i,j} = \sum_{k = 1}^m{ J_{k,i} J_{k,j}}  + \sum_{k = 1}^m {r_k * \partial{r_k}/{\partial p_i \partial p_j}}$$
Let us use $D_{i,j}$ for the elements of the second term of this expression.
What is generally called **Newton's method** for function minimization tries
to set the gradient to zero (to find a stationary point of the function $S(p)$).
This leads to **Newton's equation**
$$ H \delta = -g $$
Given a set of parameters $p$, we solve this equation for $\delta$, adjust $p$
to $p + \delta$ and iterate, hopefully to converge on a solution. 
Applying this to a sum of squares problem gives
$$ H \delta = (J'J + D) \delta = - J' r$$
In this expression, only the elements of $D$ have second partial derivatives. 
Gauss, attempting to model planetary orbits, had small residuals, and noted
that these multiplied the second partial derivatives of $r$, so he approximated
$$ H \approx J' J$$
by assuming $D \approx 0$. This results in the Gauss-Newton method where we
solve 
$$ J' J \delta = - J' r$$
though we can avoid some loss of accuracy by NOT forming the inner product matrix
$J' J$ and solving the linear least squares matrix problem 
$$ J \delta \approx -r $$
by one of several matrix decomposition methods.

@@

The usual suggestion is that $D$ be replaced by a multiple of the unit matrix or else
a multiple of the diagonal part of $J' J$. In low precision, some elements of 
$J' J$ could underflow to zero (@jn77ima), and a linear combination of both
choices is an effective compromise. Various choices for
$D$, as well as a possible line search along the
direction $\delta$ rather than a unit step (@Hartley1961), 
give rise to several variant
algorithms. "Marquardt's method" is a family of methods. Fortunately, most choices
work well.

@@

Let us specify in R the three model formulas and set some starting
values for parameters. These starting points are NOT equivalent and
are deliberately crude choices. Workers performing many calculations 
of a similar nature should try to provide good starting points to reduce
computation time and avoid finding a false solution.

@@

Using a formula specification was a principal advantage made
with `nls()` when it became available in S sometime in the 1980s. It uses a
Gauss-Newton (i.e., unstabilized) iteration with a step reduction line
search. This works very efficiently as long as $J$ is not ill-conditioned.
Below we see `nls()` does poorly on the example problem. To save page space,
we use 1-line result display functions from package `nlsr`, namely `pnls()` 
and `pshort()`.

@@

```{r ex02minpack}
unlm1<-try(nlsLM(formula=frmu, start=stu1, data=weeddf))
pnls(unlm1)  # Short form of output
snlm1<-try(nlsLM(formula=frms, start=sts1, data=weeddf))
pnls(snlm1)
tnlm1<-try(nlsLM(formula=frmt, start=stt1, data=weeddf))
pnls(tnlm1) # short form to give sum of squares, else use summary(tnlm1)
```

@@

```{r ex02gslnls}
library(gslnls)
ugslnls1<-try(gsl_nls(fn = frmu, data = weeddf,  start = stu1))
pnls(ugslnls1) # to get sum of squares
sgslnls1<-try(gsl_nls(fn = frms, data = weeddf,  start = sts1))
pnls(sgslnls1) # Use summary() to get display
tgslnls1<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1))
pnls(tgslnls1) 
```

### Comparison notes for formula-setup solutions

@@

We illustrate how to solve nonlinear least squares problems using
a function to define the residual. Note that `gsl_nls()` requires
a vector `y` that is the expected value of what we have called the
residual, but in this case is actually the model. `gsl_nls` uses 
a numerical approximation for the Jacobian if the argument `jac` 
is missing. Note function `nlsr::pnlslm()` for a 1-line display of
the results of `minpack.lm::nls.lm()`.

```{r exhobbsfn, eval=TRUE}
hobnlfb<-nlfb(start=stu1, resfn=hobbs.res, jacfn=hobbs.jac)
pshort(hobnlfb) # use print(hobnlfb) for more detail
hobnlm<-nls.lm(par=stu1, fn=hobbs.res, jac=hobbs.jac)
pnlslm(hobnlm)  
hobgsln<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12))
pnls(hobgsln)
hobgsl<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12), jac=hobbs.jac)
pnls(hobgsl) # using analytic Jacobian
```

@@

As mentioned, the output of `nls()` is an object of class "nls" which has a quite rich structure
described in the manual file or revealed by applying the `str()` function to the
result of `nls()`.

The complexity of this object is a challenge to users. Let us use 
`result` as the returned object from `nls()`, `minpack.lm::nlsLM()` or
`gslnls::gsl_nls()`. For example, the `data` return element is an R symbol. To
actually access the data from this element, we need to use the syntax:
```
eval(parse(text=result$data))
```
However, if the call is made with `model=TRUE`, then there is a returned element
`model` which contains the data, and we can list its contents using:
```
ls(result$model)
```
and if there is an element called `xdata`, then it can be accessed as
`result$model$xdata`.

@@

Using the "gradient" attribute of the output of the Jacobian function to hold
the Jacobian matrix lets us embed this in the **residual** function as well,
so that the call to `nlsr::nlfb()` can be made with the same name used for 
both residual and Jacobian function arguments. This programming trick saves a 
lot of trouble for the package developer, but it can be a nuisance for users 
trying to understand the code.

@@

The `nls()` function, after
computing the "right hand side" or `rhs` of the residual, checks to see if the
"gradient" attribute is defined, otherwise using `numericDeriv()` to compute a 
Jacobian into that attribute. This code is within the `nlsModel()` or
`nlsModel.plinear()` functions. The use of analytic Jacobians
almost certainly contributes to the good performance of `nls()` on `selfStart` 
models.

@@

In considering `selfStart` models, we noted that the base-R function `SSlogis`
is intended to solve problem **Logistic3T** above.  When this function is used
via `getInitial()` to  find
starting values, it actually calls `nls()` with the 'plinear' algorithm 
and finds a (full) solution. It then passes the solution coefficients to the default 
algorithm unnecessarily. Moreover, the implicit double call is, in our view, prone to creating 
errors in code maintenance. 
To provide simpler starting parameters, the function `SSlogisJN`
is now part of package `nlsr`, but is most useful for `nls()`.

Users may also want to profit from the Jacobian code of `selfStart` models 
but supply explicit starting values other than those suggested by `getInitial()`.
This does not appear to be possible with `nls()`. `nlsr::nlxb()` always requires 
starting parameters, and can either use `getInitial()` to find them from the `selfStart`
model or provide explicit values, but the `selfStart` model is used in the formula 
for the `nlxb()` call.

@@

We are also surprised that the 
analytic expressions for the Jacobian ("gradient") in the `SSLogis` function 
and others save quantities in "hidden"
variables, i.e., with names preceded by ".". 
These are then not displayed by the `ls()`
command, making them difficult to access by users who may wish to create
their own selfStart model via copy and edit.
Interactive tools, such as "visual fitting" (@nash1996nonlinear) might be
worth considering as another way to find starting parameters, but we know of no
R capability of this type.

@@

For many problems we know that parameters cannot be bigger or smaller than some
externally known limits. Such limits should be built into models, but there 
are some important details for using the tools in R.

@@

```{r ex10, echo=TRUE} 
# Start MUST be feasible i.e. on or within bounds
anlshob1b <- nls(frms, start=sts1, data=weeddf, lower=c(0,0,0),
             upper=c(2,6,3), algorithm='port')
pnls(anlshob1b) #  check the answer (short form)
# nlsLM seems NOT to work with bounds in this example
anlsLM1b <- nlsLM(frms, start=sts1, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
pnls(anlsLM1b)
# also no warning if starting out of bounds, but gets good answer!!
st4<-c(c1=4, c2=4, c3=4)
anlsLMob <- nlsLM(frms, start=st4, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
pnls(anlsLMob)
# Try nlsr::nlxb()
anlx1b <- nlxb(frms, start=sts1, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
pshort(anlx1b)
```

@@

```{r ex10m, echo=TRUE}
# Hobbsmaskx.R -- masks with formula specification of problem
require(nlsr); require(minpack.lm); traceval<-FALSE
stu <- c(b1=200, b2=50, b3=0.3) # a default starting vector (named!)
sts <- c(c1=2, c2=5, c3=3) # a default scaled starting vector (named!)
# fix first parameter
anxbmsk1 <- try(nlxb(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
			upper=c(200, 60, 3), trace=traceval))
print(anxbmsk1)
anlM1 <- try(nlsLM(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
			upper=c(200, 60, 3), trace=traceval))
pnls(anlM1)
anlsmsk1 <- try(nls(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
		upper=c(200, 60, 3),  algorithm="port", trace=traceval))
pnls(anlsmsk1)
# Hobbs scaled problem with bounds, formula specification
anlxmsks1 <- nlxb(frms, start=sts, data=weeddf, lower=c(2,0,0),
                  upper=c(2,6,30))
print(anlxmsks1)
anlshmsk1 <- nls(frms, start=sts, trace=traceval, data=weeddf, lower=c(2,0,0),
             upper=c(2,6,30), algorithm='port')
pnls(anlshmsk1)
anlsLMmsks1 <- nlsLM(frms, start=sts, data=weeddf, lower=c(2,0,0),
                 upper=c(2,6,30))
pnls(anlsLMmsks1)

@@

`nls()` can be called without specifying the `data` argument. In this case, it will
search in the available environments (i.e., workspaces) for suitable data objects. 
We do NOT like this approach, but it is "the R way". R allows users to leave many 
objects in the default (`.GlobalEnv`) workspace. Moreover, users have to actively 
suppress saving this workspace (`.RData`) on exit, otherwise any such file in 
the path will be loaded on startup. R users in our
acquaintance avoid saving the workspace because of lurking data and
functions that may cause unwanted results.

@@

`na.action` is an argument to the `nls()` function, but it does not appear 
obviously in the source code, often being handled behind the scenes after 
referencing the option `na.action`. This feature also changes the data 
supplied to our nonlinear least squares solver. 
A useful, but possibly dated, description is given in:
*https://stats.idre.ucla.edu/r/faq/how-does-r-handle-missing-values/*.
The typical default action, which can be seen by using the command
`getOption("na.action")`
is `na.omit`. This option removes from computations any observations
containing missing values (i.e. any row of a data frame containing an NA). 
`na.exclude` does much of the same for solver computations, but keeps the rows with 
NA elements so that predictions are in the correct row position. We recommend that 
workers actually test output to verify the behaviour is as wanted.
See *https://stats.stackexchange.com/questions/492955/should-i-use-na-omit-or-na-exclude-in-a-linear-model-in-r*.
As with `subset`, our concern with `na.action` is that users may be unaware of the
effects of an option they may not even be aware has been set. Should `na.fail` be the default?

@@

While fixed weights may be useful, there are many problems for which
we want weights that are determined at least partially from the model 
parameters, for
example, a measure of the standard deviation of observations.
Such dynamic weighting situations are discussed in the vignette 
"Introduction to nlsr" of package `nlsr` in section 
*Weights that are functions of the model parameters*.
`minpack.lm` offers a function `wfct()` to 
facilitate such weighting. Care is advised
in applying such ideas.

@@

The variable projection method (@Golub1973, @OlearyRust13) is 
generally much more effective than general approaches in finding good solutions 
to nonlinear least squares 
problems when some of the parameters appear linearly. In our logistic examples, 
the asymptote parameters are an illustration. However, identifying which parameters
are linear and communicating this information to estimating functions 
is not a trivial task. `nls()` has an option `algorithm="plinear"` that allows some partially linear
models to be solved. The other tools, as far as we are aware, do not offer any
such capability. The `nlstac` package uses a different algorithm for similar goals.

@@

Maintainers of packages need suitable tests and use-case examples in order

- to ensure packages work properly, in particular, giving
results comparable to or better than the functions they are to replace. 
- to test individual solver functions to ensure they work across the range of 
  calling mechanisms, that is, different ways of supplying inputs to the solver(s);
- to pose "silly" inputs to see if these bad inputs are caught by the programs.
